### 2015-01-16 partialtotal.tex
Initial commit.

\documentclass[10pt,a4paper]{article}
\usepackage{a4wide}

\usepackage{ifthen}

\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\DeclareUnicodeCharacter{8656}{\ensuremath{\Leftarrow}}
\DeclareUnicodeCharacter{8614}{\ensuremath{{\mapsto}}}
\DeclareUnicodeCharacter{8871}{\ensuremath{\models}}
\DeclareUnicodeCharacter{963}{\ensuremath{\sigma}}
\DeclareUnicodeCharacter{956}{\ensuremath{\mu}}
\DeclareUnicodeCharacter{957}{\ensuremath{\nu}}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{johnproof}
\usepackage{verbatim}

% THEOREMS
\theoremstyle{theorem}
\newtheorem{todo}{Todo}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}
\newtheorem*{conjecture}{Conjecture}
\newtheorem{lem}[theorem]{Lemma}
\newtheorem{cor}[theorem]{Corollary}
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{rem}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}

\title{\vspace{-60pt} Partial and total correctness as greatest and least fixpoints}
\author{John Wickerson}
\date{25th November 2009}

\newcommand{\eqdef}{=_{\mathrm{def}}}
\newcommand{\prov}[1][]{\ifthenelse{\equal{#1}{}}{\vdash}{\vdash_{\mbox{\sf\scriptsize #1}}}}
\newcommand{\seqspec}[3]{\{#1\}\,#2\,\{#3\}}
\newcommand{\totalspec}[3]{[#1]\,#2\,[#3]}

\renewcommand{\parallel}{\mathbin{\mathsf{ll}}}
\newcommand{\semicolon}{\mathbin{\mathbf{;}}}
\newcommand{\Skip}{\mathop{\mathtt{skip}}}

\begin{document}
\maketitle
\thispagestyle{empty}

\begin{abstract}
This short note explains how partial and total correctness triples can be characterised as the greatest and least fixpoints, respectively, of the same function.
\end{abstract}

Suppose we have a small-step transition relation $→$ between configurations. Configurations comprise a command $C$ and a state σ. Given an initial configuration $(C_0,σ_0)$, we can produce a transition system corresponding to all the possible computations that can arise. A simple one is as follows:
\[
\begin{array}{c}
(\texttt{x++} \parallel \texttt{y++},\{{\tt x}↦0,{\tt y}↦2\}) \\
\swarrow \hspace{40pt} \searrow \\
({\tt skip} \parallel \texttt{y++},\{{\tt x}↦1,{\tt y}↦2\}) \qquad \qquad (\texttt{x++} \parallel {\tt skip},\{{\tt x}↦0,{\tt y}↦3\}) \\
\searrow \hspace{40pt} \swarrow \\
({\tt skip}\parallel {\tt skip},\{{\tt x}↦1,{\tt y}↦3\}) \\
↓\\
({\tt skip},\{{\tt x}↦1,{\tt y}↦3\})
\end{array}
\]

We can use modal-μ calculus to describe such transition systems. We shall consider modal-μ assertions with the following, slightly reduced, syntax:
\[
p ::= S \mid \top \mid \bot \mid p ⇒ p \mid p ∧ p \mid \square p \mid \diamond p \mid μX\ldotp p \mid νX\ldotp p \mid X
\]
where $S$ is a set of configurations and $μX\ldotp p$ and $νX\ldotp p$ are the least and greatest assertions such that $X=p$, where $p$ contains $X$. The satisfaction relation $(C,σ) ⊧ p$ is defined as follows:
\[
\begin{array}{lcl}
(C,σ)⊧S & ⇔ & (C,σ)\in S \\
(C,σ)⊧\top & ⇔ & \mathit{true} \\
(C,σ)⊧\bot & ⇔ & \mathit{false} \\
(C,σ)⊧p_1⇒p_2 & ⇔ & (C,σ)⊧p_1 ⇒ (C,σ)⊧p_2 \\
(C,σ)⊧p_1∧p_2 & ⇔ & (C,σ)⊧p_1 ∧ (C,σ)⊧p_2 \\
(C,σ)⊧\square p & ⇔ & ∀C',σ'\ldotp (C,σ)→(C',σ') ⇒ (C',σ')⊧p \\
(C,σ)⊧\diamond p & ⇔ & ∃C',σ'\ldotp (C,σ)→(C',σ') ∧ (C',σ')⊧p \\
(C,σ)⊧μX\ldotp p & ⇔ & (C,σ)\in \bigcap\{S\mid p[S/X] ⊆ S\} \\
(C,σ)⊧νX\ldotp p & ⇔ & (C,σ)\in \bigcup\{S\mid S ⊆ p[S/X]\}
\end{array}
\]
Let $P$ and $Q$ range over sets of states. Understand $(C,P)$ as a shorthand for $\{(C,σ)\mid σ\in P\}$, and let $(C,P)⊧p$ mean $∀σ\in P\ldotp (C,σ)⊧p$.

\begin{definition}[Termination] A configuration $(C,σ)$ is \emph{stuck} if no further reductions are possible. Moreover, the configuration is \emph{terminal} when $C={\tt skip}$.
\end{definition}

\begin{definition}[Partial correctness] $\seqspec{P}C{Q}$ means that whenever $C$ is executed from an initial state in $P$, then in every terminal configuration it reaches, the state is in $Q$.
\end{definition}

\begin{definition}[Total correctness] $\totalspec{P}C{Q}$ means that whenever $C$ is executed from an initial state in $P$, then it reaches a terminal configuration, and in every terminal configuration it reaches, the state is in $Q$.
\end{definition}

\begin{conjecture}Partial correctness and total correctness are characterised by the following modal-μ assertions.
\[
\begin{array}{rcl}
\seqspec{P}C{Q} & ⇔ & (C,P) ⊧ νX\ldotp \square X ∧ (\square\bot ⇒ ({\tt skip},Q)) \\
\totalspec{P}C{Q} & ⇔ & (C,P) ⊧ μX\ldotp \square X ∧ (\square\bot ⇒ ({\tt skip},Q)) 
\end{array}
\]
\end{conjecture}
\begin{proof}To do.\end{proof}

\end{document}

### 2015-01-16 paper.tex
Initial commit (again).

\documentclass{llncs}

\pdfpagesattr{/CropBox [92 72 523 738]} % LNCS page: 152x235 mm

\title{Partial and total correctness \\ as greatest and least fixed points}
\author{John Wickerson}
\institute{Imperial College London}
\date{\today}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{../../mathpartir}

\usepackage{../../MathUnicode}
\usepackage{mathtools}
\usepackage{bm}

\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{decorations.pathreplacing}

\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}

\newcommand{\prov}[1][]{\ifthenelse{\equal{#1}{}}{\vdash}{\vdash_{\mbox{\sf\scriptsize #1}}}}
\newcommand{\seqspec}[3]{\{#1\}\,#2\,\{#3\}}
\newcommand\boldleftbracket{\phantom{\{}\mathllap{\bm{[}}}
\newcommand\boldrightbracket{\mathrlap{\bm{]}}\phantom{\}}}
\newcommand{\totalspec}[3]{\boldleftbracket #1\boldrightbracket\,#2\,\boldleftbracket#3\boldrightbracket}

\renewcommand{\parallel}{\mathbin{\mathsf{ll}}}
\newcommand{\semicolon}{\mathbin{\mathbf{;}}}
\newcommand{\Skip}{\mathop{\mathtt{skip}}}

\newcommand\squigarrow[1]{
\tikz[rotate=#1]
\draw [->, line join=round, decorate, 
  decoration={zigzag, segment length=4, amplitude=.9,
  post=lineto, post length=2pt}] (0,0) -- (0.3,0);
}
\newcommand\trans{\rightsquigarrow}
\newcommand\Config{\mathsf{Config}}
\newcommand\State{\mathsf{State}}
\newcommand\Cmd{\mathsf{Cmd}}
\newcommand\Next{\mathit{next}}
\newcommand\SafeOne{\varphi}
\newcommand\terminates{\mathit{always\text{-}terminates}}
\newcommand\dom{\mathrm{dom}}
\newcommand\length{\mathrm{len}}
\newcommand\fst{\mathrm{fst}}
\newcommand\snd{\mathrm{snd}}
\newcommand\Safe{\mathit{safe}}
\newcommand\tracesfrom{\mathit{traces}}
\newcommand\BOX{\mathop{\Box}}
\newcommand\DIAMOND{\mathop{\Diamond}}
\newcommand\pow{\mathcal{P}}
\newcommand\nat{\mathbb{N}}
\newcommand\undef{\bot}
\newcommand\stuck[1]{\mathit{stuck}#1}
\newcommand\config[2]{\langle #1,#2\rangle}
\newcommand\infinite{\mathit{infinite}}
\newcommand\finite{\mathit{finite}}

\begin{document}

\maketitle

\thispagestyle{plain}
\pagestyle{plain}

\begin{abstract}
I explain how the partial correctness interpretation of a Hoare triple can be characterised as the greatest fixed point of a function, and how the total correctness interpretation can be seen as the least fixed point of the same function. In the latter case, I provide a necessary and sufficient condition for the characterisation to be accurate: that the programming language admits no infinite branching.
\end{abstract}

\section{Introduction}

In the context of Hoare Logic~\cite{hoare69}, a program satisfies a \emph{total correctness} specification when each of its traces from state that satisfies the given precondition reaches a terminal state satisfying the given postcondition within a finite number of steps. That program satisfies a \emph{partial correctness} specification when each of its traces either meets the requirement above or is infinite.

Meanwhile, in the context of order theory, an object is in the \emph{least fixed point} of a continuous function if, roughly speaking, it can be shown meet some condition within a finite number of iterations. That object is in the \emph{greatest fixed point} if it is in the least fixed point or is, in some sense, infinite.

The descriptions above have been designed to suggest a connection between total/partial correctness and least/greatest fixed points. This paper seeks to state this connection precisely, and investigate conditions under which it holds.

Section 3 provides characterisations of partial and total correctness, that differ only in that the former takes a function's greatest fixed point where the latter takes its least. Section 4 describes a condition that is necessary for the fixed point characterisation of total correctness to be accurate: namely, that there is no infinite branching.

\section{Preliminaries}

We assume a small-step transition relation $\trans$ between configurations $C=\config{c}{σ}$, which comprise a command $c ∈ \Cmd$ and a state $σ ∈ \State$. Let $\Config = \Cmd × \State$ be the set of all configurations. We use the abbreviation $\Next(C) \eqdef \{C'\ldotp C \rightsquigarrow C'\}$, and we write $\stuck{(C)}$ if $C$ admits no further transitions. The functions $\fst$ and $\snd$ serve to project the components of a pair.

\subsubsection{Modal-$μ$ calculus} We employ the following constructions from the modal-$μ$ calculus to describe properties of our transition relation. In the following, we suppose that $p ∈ \pow(\Config)$ and that $\varphi:\pow(\Config) → \pow(\Config)$ is a monotone function.
\[\begin{array}{r@{~~}c@{~~}l}
\BOX p & = & \{C \mid ∀C'∈\Next(C)\ldotp C' ∈ p\} \\
\DIAMOND p & = & \{C \mid ∃C'∈\Next(C)\ldotp C' ∈ p\} \\
μX\ldotp \varphi(X) & = & \bigcap\{S\mid \varphi(S) ⊆ S\} \\
νX\ldotp \varphi(X) & = & \bigcup\{S\mid S ⊆ \varphi(S)\}
\end{array}\]

\subsubsection{Possibly-infinite sequences} A \emph{possibly-infinite $X$-sequence} is a partial function $π : \nat → X ∪ \{\undef\}$
whose domain of definition is either the entirety of $\nat$ or an initial subset thereof. In the latter case, we define $\length(π) = j+1$ when $j$ is the greatest natural in $π$'s domain.

\subsubsection{Traces} A trace is a possibly-infinite sequence of configurations, successively related by $\trans$. The set of traces beginning from a configuration $C$, written $\tracesfrom(C)$, comprises those sequences $\pi$ for which $π_0 = C$ and for all $i$:
\begin{eqnarray*}
π_{i+1} = \undef & ~~\text{iff}~~ & \text{$π_i = \undef$ or $\stuck{(π_i)}$} \\
π_{i+1} ∈ \Next(C) & ~~\text{iff}~~ & \text{not $\stuck{(π_i)}$}
\end{eqnarray*}
We write $π_i$ instead of the usual $π(i)$ to save space.

\subsubsection{Termination} A configuration $C$ \emph{always terminates} if every trace from $C$ reaches a terminal configuration.
\[
\begin{array}{r@{~~}c@{~~}l}
  C ∈ \terminates &\eqdef& ∀π ∈ \tracesfrom(C)．∃j>0．\length(π) = j
\end{array}
\]

\subsubsection{Safe configurations} If $Q ∈ \pow(\State)$ is a postcondition, we say that a configuration $C$ is \emph{safe for $Q$}, written $C ∈ \Safe_Q$, if whenever it reaches a terminal configuration, the state is in $Q$.
\begin{eqnarray*}
C ∈ \Safe_Q &\eqdef& ∀c',σ'．((C \trans^* \config{c'}{σ'}) ∧ \stuck{\config{c'}{σ'}}) ⇒ σ' ∈ Q
\end{eqnarray*}

\subsubsection{Partial and total correctness} Suppose $P,Q ∈ \pow(\State)$ and $c ∈ \Cmd$. We write $\seqspec{P}c{Q}$ to mean that whenever $c$ is executed from a state in $P$, then whenever it reaches a terminal configuration, the state is in $Q$. We write $\totalspec{P}c{Q}$ to mean that whenever $c$ is executed from a state in $P$, then it reaches a terminal configuration, and whenever it reaches a terminal configuration, the state is in $Q$.
\[
\begin{array}{r@{~~}c@{~~}l}
\seqspec{P}c{Q} &\eqdef& ∀σ ∈ P．\config{c}{σ} ∈ \Safe_Q \\
\totalspec{P}c{Q} &\eqdef& ∀σ ∈ P．\config{c}{σ} ∈ \terminates ∩ \Safe_Q
\end{array}
\]

\section{Main result}

\begin{theorem} Assuming finite branching -- that is, that $\Next(C)$ is finite for every $C$ -- partial correctness and total correctness can be characterised as follows.
\begin{eqnarray}
\label{thm1_pc} \seqspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ νX\ldotp \SafeOne_Q(X) \\
\label{thm1_tc} \totalspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X) 
\end{eqnarray}
where $\SafeOne_Q(X)$ is the set of configurations for which the state is in $Q$ if the configuration is terminal, and
every next configuration is in $X$:
\[
\begin{array}{r@{~~}c@{~~}l}
\SafeOne_Q(X) &\eqdef& \{\config{c}{σ}\mid  \stuck{\config{c}{σ}} ⇒ σ ∈ Q\} ∩ \Box X.
\end{array}
\]
\end{theorem}

\begin{proof}
We begin with the fixed point characterisation of partial correctness. To prove \eqref{thm1_pc}, it suffices to prove
\begin{eqnarray*}
\Safe_Q &=& νX\ldotp \SafeOne_Q(X).
\end{eqnarray*}
For the $(⊆)$ direction, we show that $\Safe_Q$ is a post-fixed point (that is, $\Safe_Q ⊆ \SafeOne_Q (\Safe_Q)$), and immediately deduce that it is below the \emph{greatest} post-fixed point.

In showing the $(⊇)$ direction, we observe that since $\SafeOne_Q$ preserves greatest lower bounds, the greatest fixed point can be constructed as the intersection of a series of approximants:
\begin{eqnarray*}
νX\ldotp \SafeOne_Q(X) &=& \bigcap_{n∈\nat}\left(\SafeOne_Q^n(\Config)\right).
\end{eqnarray*}

The intuition is that the $k$th approximant only contains configurations for which every trace that terminates in fewer than $k$ steps satisfies the postcondition when it does so.

We now turn to the fixed point characterisation of total correctness. To prove \eqref{thm1_tc}, it suffices to prove
\[
∀C ∈ \SafeOne_Q^k (\Config). ∀π ∈ \tracesfrom(C)．∀j<k．\length(π) = j+1 ⇒ \snd (π_j) ∈ Q
\] 
which can be dispatched via mathematical induction on $k$.

It suffices to show that
\begin{eqnarray*}
\Safe_Q ∩ \terminates &=& μX\ldotp \SafeOne_Q(X).
\end{eqnarray*}
For the $(⊇)$ direction, we show that $\Safe_Q$ is a pre-fixed point (that is, $\SafeOne_Q (\Safe_Q) ⊆ \Safe_Q$), and immediately deduce that it is above the \emph{least} pre-fixed point.

In showing the $(⊆)$ direction, we observe that since $\SafeOne_Q$ preserves least upper bounds (this relies on the finitary nature of $\Next(C)$), the least fixed point can be constructed as the union of a series of approximants:
\begin{eqnarray*}
μX\ldotp \SafeOne_Q(X) &=& \bigcup_{n∈\nat}\left(\SafeOne_Q^n(∅)\right).
\end{eqnarray*}
The intuition is that the $k$th approximant contains all configurations for which every trace terminates in fewer than $k$ steps and satisfies the postcondition when it does so.

It remains then to show
\[
∀C．(∀π ∈ \tracesfrom(C)． ∃j < k．\length(π) = j+1 ∧ \snd (π_j) ∈ Q) ⇒ C ∈ \SafeOne_Q^{k+1}(\emptyset)
\]
which can be dispatched via mathematical induction on $k$.
\end{proof}

\section{Proof mechanisation}

The proof presented in the previous section has been formalised in the Isabelle theorem prover, with the exception of the following lemma, which has currently only been proved `by hand'. It states that in the absence of infinite branching, every command that always terminates has a longest trace.
\begin{lemma}
If $\Next(C)$ is finite for all $C$, and $C_0 ∈ \terminates$, then there exists an upper bound $M$ for which $∀π ∈ \tracesfrom(C_0)．∃j≤M．\length(π) = j$.
\end{lemma}
The lemma holds because $\tracesfrom(C_0)$ is a finite set; this observation in turn rests on K\H{o}nig's infinity lemma.

\section{On infinite branching}

If the $\trans$ relation allows infinite branching -- that is, if a configuration can have infinitely many next configurations to choose from -- then the least fixed point calculation does not coincide with total correctness. The technical reason for the failure is that $\SafeOne_Q$ no longer preserves least upper bounds. Intuitively, the failure can be explained by the following counterexample.
\[
\begin{tikzpicture}
\node[draw] (c0) at (3,0) {$C_0$};
\node[draw] (c21) at (1,1) {$C_1$};
\node[draw] (c22) at (1,2) {\phantom{$C_0$}};
\node[draw] (c31) at (2,1) {$C_2$};
\node[draw] (c32) at (2,2) {\phantom{$C_0$}};
\node[draw] (c33) at (2,3) {\phantom{$C_0$}};
\node[draw] (c11) at (3,1) {$C_3$};
\node[draw] (c12) at (3,2) {\phantom{$C_0$}};
\node[draw] (c13) at (3,3) {\phantom{$C_0$}};
\node[draw] (c14) at (3,4) {\phantom{$C_0$}};
\node             at (4,1) {$\dots$};
\node[draw] (c41) at (5,1) {$C_k$};
\node[draw] (c42) at (5,2) {\phantom{$C_0$}};
\node       (c43) at (5,3) {$\vdots$};
\node[draw] (c44) at (5,4) {\phantom{$C_0$}};
\node             at (6,1) {$\dots$};
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c21);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c21) -- (c22);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c31);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c31) -- (c32);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c32) -- (c33);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c11);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c11) -- (c12);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c12) -- (c13);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c13) -- (c14);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c41);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c41) -- (c42);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c42) -- (c43);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c43) -- (c44);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] [decorate,decoration={brace,amplitude=10pt}]
([xshift=5mm]c44.north) -- ([xshift=5mm]c42.south) node [black,midway,xshift=6mm] {$k$};
\end{tikzpicture}
\]
The $k$th approximant contains configurations whose traces all terminate in fewer than $k$ steps. No approximant can contain configuration $C_0$, since there is no bound $k$ on the length of its traces. Yet this configuration will be admitted by a total correctness specification whose postcondition is $\mathit{true}$, since it is the case that each of its traces terminates.

Therefore, in the presence of infinite branching, the fixed point characterisations of total correctness must be weakened to an implication. Preservation of greatest lower bounds is unaffected by infinite branching, so the characterisation of partial correctness remains intact.

\begin{eqnarray*}
\seqspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ νX\ldotp \SafeOne_Q(X) \\
\totalspec{P}c{Q} & ~⇐~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X) 
\end{eqnarray*}

\section{Discussion}

Recent work on various Hoare logics for concurrency have used (mildly disguised) greatest fixed point calculations to obtain a partial correctness semantics. See, for example, \cite[Definition~3.2]{vafeiadis11} and \cite[Definition~25]{dinsdale-young+13}. The result in this paper could prove helpful when those program logics are extended to total correctness.

Jacobs and Gries have proposed \emph{general correctness} as a way to unify partial and total correctness~\cite{jacobs+85}. It would be interesting to investigate whether general correctness can be also characterised as a fixed point calculation.

The idea of using the least and greatest fixed point of the same function has been previously exploited by Paulson~\cite[\S3]{paulson97a}, who obtains the set of finite lists from a function's least fixed point, and the set of possibly-infinite (`lazy') lists from its greatest fixed point.

\paragraph{Acknowledgements} This work was supported by EPSRC grant EP/K011499/1. I would like to thank Andreas Lochbihler for several helpful discussions.

\bibliographystyle{abbrv}
\bibliography{../../John}

\end{document}

### 2015-01-20 paper.tex
First working commit
+

@@
 
+% Define a version of \paragraph whose heading runs straight into the
+% text that follows. {
+\makeatletter
+\usepackage{suffix}
+\renewcommand\paragraph[1]{%
+\@startsection{paragraph}{3}{\z@}%
+{-8\p@ \@plus -4\p@ \@minus -4\p@}%
+{-1em \@plus -0.22em \@minus -0.1em}%
+{\normalfont\normalsize\bfseries\boldmath}{#1}}
+\renewcommand\subparagraph[1]{%
+\@startsection{paragraph}{3}{\z@}%
+{-8\p@ \@plus -4\p@ \@minus -4\p@}%
+{-1em \@plus -0.22em \@minus -0.1em}%
+{\normalfont\normalsize\itshape}{#1}}
+\WithSuffix\newcommand\paragraph*[1]{%
+\@startsection{paragraph}{3}{\z@}%
+{-8\p@ \@plus -4\p@ \@minus -4\p@}%
+{0pt}%
+{\normalfont\normalsize\bfseries\boldmath}{#1} \hspace*{0pt}}
+\makeatother
+% }
+
 \title{Partial and total correctness \\ as greatest and least fixed points}
@@
 \institute{Imperial College London}
-\date{\today}
 
@@
 \usepackage{amssymb}
-\usepackage{../../mathpartir}
 
-\usepackage{../../MathUnicode}
+\usepackage{MathUnicode}
+\usepackage{mathpartir}
 \usepackage{mathtools}
 \usepackage{bm}
+
+\renewcommand\rmdefault{ptm}
 
@@
 \begin{abstract}
-I explain how the partial correctness interpretation of a Hoare triple can be characterised as the greatest fixed point of a function, and how the total correctness interpretation can be seen as the least fixed point of the same function.
-In the latter case, I provide a necessary and sufficient condition for the characterisation to be accurate: that the programming language admits no infinite branching.
+This paper studies Hoare triples in the context of any programming
+language specified by a small-step, possibly non-deterministic,
+operational semantics. We explain how the partial correctness
+interpretation of the triple can be characterised as the greatest
+fixed point of a function, and how the total correctness
+interpretation can be seen as the least fixed point of that very same
+function. In the latter case, we provide a necessary and sufficient
+condition for the characterisation to be accurate: that the
+programming language admits no infinite branching.
 \end{abstract}
@@
 
-In the context of Hoare Logic~\cite{hoare69}, a program satisfies a \emph{total correctness} specification when each of its traces from state that satisfies the given precondition reaches a terminal state satisfying the given postcondition within a finite number of steps.
-That program satisfies a \emph{partial correctness} specification when each of its traces either meets the requirement above or is infinite.
+In the context of Hoare Logic~\cite{hoare69}, a (possibly
+non-deterministic, possibly non-terminating) program satisfies a
+\emph{total correctness} specification when each of its traces from a
+state that satisfies the given precondition reaches a terminal state
+satisfying the given postcondition within a finite number of execution
+steps. That program satisfies a \emph{partial correctness}
+specification when each of its traces either meets the requirement
+above or is infinite.
 
-Meanwhile, in the context of order theory, an object is in the \emph{least fixed point} of a continuous function if, roughly speaking, it can be shown meet some condition within a finite number of iterations.
-That object is in the \emph{greatest fixed point} if it is in the least fixed point or is, in some sense, infinite.
+Meanwhile, in the context of order theory, an
+object is in the \emph{least fixed point} of a continuous function if
+(roughly speaking) it can be shown meet some condition within a finite
+number of iterations. That object is in the \emph{greatest fixed
+point} if it is in the least fixed point or is, in some sense,
+infinite.
 
-The descriptions above have been designed to suggest a connection between total/partial correctness and least/greatest fixed points.
-This paper seeks to state this connection precisely, and investigate conditions under which it holds.
+The descriptions above have been deliberately crafted to emphasise a
+connection between total/partial correctness and least/greatest fixed
+points. This paper seeks to state this connection precisely, and
+investigate conditions under which it holds.
 
-Section 3 provides characterisations of partial and total correctness, that differ only in that the former takes a function's greatest fixed point where the latter takes its least.
-Section 4 describes a condition that is necessary for the fixed point characterisation of total correctness to be accurate: namely, that there is no infinite branching.
+Section~\ref{sec:result} provides characterisations of partial and
+total correctness, that differ only in that the former takes a
+function's greatest fixed point where the latter takes its least, and
+Section~\ref{sec:isabelle} describes how this characterisation is
+proved correct using the Isabelle theorem
+prover. Section~\ref{sec:infinite_branching} describes a condition
+that is necessary for the fixed point characterisation of \emph{total}
+correctness to be accurate; namely, that there is no infinite
+branching. We begin, in Section~\ref{sec:preliminaries}, by
+establishing some preliminary definitions.
 
 \section{Preliminaries}
+\label{sec:preliminaries}
 
-We assume a small-step transition relation $\trans$ between configurations $C=\config{c}{σ}$, which comprise a command $c ∈ \Cmd$ and a state $σ ∈ \State$.
-Let $\Config = \Cmd × \State$ be the set of all configurations.
-We use the abbreviation $\Next(C) \eqdef \{C'\ldotp C \rightsquigarrow C'\}$, and we write $\stuck{(C)}$ if $C$ admits no further transitions.
-The functions $\fst$ and $\snd$ serve to project the components of a pair.
+We assume a small-step transition relation $\trans$ between
+configurations $C=\config{c}{σ}$, which comprise a command $c ∈ \Cmd$
+and a state $σ ∈ \State$. We impose no constraints on the forms taken
+by commands and states. Let $\Config = \Cmd × \State$ be the set of
+all configurations. We use the abbreviation
+$\Next(C) \eqdef \{C'\ldotp C \rightsquigarrow C'\}$ for the set of
+configurations immediately reachable from $C$, and we write
+$\stuck{(C)}$ if $C$ admits no further transitions. The functions
+$\fst$ and $\snd$ serve to project the components of a pair.
 
-\subsubsection{Modal-$μ$ calculus} We employ the following constructions from the modal-$μ$ calculus to describe properties of our transition relation.
-In the following, we suppose that $p ∈ \pow(\Config)$ and that $\varphi:\pow(\Config) → \pow(\Config)$ is a monotone function.
+\paragraph{Modal-$μ$ calculus} We employ the following constructions
+from the modal-$μ$ calculus~\cite{kozen83} to describe properties of our
+transition relation. In the following, we suppose that
+$p ∈ \pow(\Config)$ and that $\varphi:\pow(\Config) → \pow(\Config)$
+is a monotone function.
 \[\begin{array}{r@{~~}c@{~~}l}
@@
 
-\subsubsection{Possibly-infinite sequences} A \emph{possibly-infinite $X$-sequence} is a partial function $π : \nat → X ∪ \{\undef\}$
-whose domain of definition is either the entirety of $\nat$ or an initial subset thereof. In the latter case, we define $\length(π) = j+1$ when $j$ is the greatest natural in $π$'s domain.
+\paragraph{Possibly-infinite sequences} Given a set $X$, a
+\emph{possibly-infinite sequence} is a partial function
+$π : \nat → X ∪ \{\undef\}$ whose domain of definition is either the
+entirety of $\nat$ or an initial subset thereof. In the latter case,
+we define $\length(π) = j+1$ when $j$ is the greatest natural in $π$'s
+domain. We shall sometimes refer to an element of a sequence by
+writing $\pi_i$ instead of $\pi(i)$.
 
-\subsubsection{Traces} A trace is a possibly-infinite sequence of configurations, successively related by $\trans$.
-The set of traces beginning from a configuration $C$, written $\tracesfrom(C)$, comprises those sequences $\pi$ for which $π_0 = C$ and for all $i$:
-\begin{eqnarray*}
-π_{i+1} = \undef & ~~\text{iff}~~ & \text{$π_i = \undef$ or $\stuck{(π_i)}$} \\
-π_{i+1} ∈ \Next(C) & ~~\text{iff}~~ & \text{not $\stuck{(π_i)}$}
-\end{eqnarray*}
-We write $π_i$ instead of the usual $π(i)$ to save space.
+\paragraph{Traces} A trace is a possibly-infinite
+sequence of configurations, successively related by $\trans$. The set of
+traces beginning from a configuration $C$, written $\tracesfrom(C)$,
+comprises those sequences $\pi$ for which $π_0 = C$ and for all $i$:
+\begin{mathpar}
+\inferrule{π_i = \undef}{π_{i+1} = \undef}
 
-\subsubsection{Termination} A configuration $C$ \emph{always terminates} if every trace from $C$ reaches a terminal configuration.
+\inferrule{\stuck{(π_i)}}{π_{i+1} = \undef}
+
+\inferrule{\neg\stuck{(π_i)}}{π_{i+1} ∈ \Next(C)}
+\end{mathpar}
+
+% \begin{eqnarray*}
+% π_{i+1} = \undef & ~~\text{iff}~~ & \text{$π_i = \undef$ or $\stuck{(π_i)}$} \\
+% π_{i+1} ∈ \Next(C) & ~~\text{iff}~~ & \text{not $\stuck{(π_i)}$}
+% \end{eqnarray*}
+
+\paragraph{Termination} A configuration $C$ \emph{always terminates}
+if every trace from $C$ reaches a terminal configuration.
 \[
@@
 
-\subsubsection{Safe configurations} If $Q ∈ \pow(\State)$ is a postcondition, we say that a configuration $C$ is \emph{safe for $Q$}, written $C ∈ \Safe_Q$, if whenever it reaches a terminal configuration, the state is in $Q$.
+\paragraph{Safe configurations} If $Q ∈ \pow(\State)$ is a
+postcondition, we say that a configuration $C$ is \emph{safe for $Q$},
+written $C ∈ \Safe_Q$, if whenever a trace starting from that
+configuration reaches a terminal configuration, the state is in $Q$.
 \begin{eqnarray*}
@@
 
-\subsubsection{Partial and total correctness} Suppose $P,Q ∈ \pow(\State)$ and $c ∈ \Cmd$.
-We write $\seqspec{P}c{Q}$ to mean that whenever $c$ is executed from a state in $P$, then whenever it reaches a terminal configuration, the state is in $Q$.
-We write $\totalspec{P}c{Q}$ to mean that whenever $c$ is executed from a state in $P$, then it reaches a terminal configuration, and whenever it reaches a terminal configuration, the state is in $Q$.
+\paragraph{Partial and total correctness} Suppose $P,Q ∈ \pow(\State)$
+and $c ∈ \Cmd$. We write $\seqspec{P}c{Q}$ to mean that whenever $c$
+is executed from a state in $P$, then whenever it reaches a terminal
+configuration, the state is in $Q$. We write $\totalspec{P}c{Q}$ to
+mean that whenever $c$ is executed from a state in $P$, then it
+reaches a terminal configuration, and whenever it reaches a terminal
+configuration, the state is in $Q$.
 \[
@@
 \section{Main result}
+\label{sec:result}
 
-\begin{theorem} Assuming finite branching -- that is, that $\Next(C)$ is finite for every $C$ -- partial correctness and total correctness can be characterised as follows.
-\begin{eqnarray}
-\label{thm1_pc} \seqspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ νX\ldotp \SafeOne_Q(X) \\
-\label{thm1_tc} \totalspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X) 
-\end{eqnarray}
-where $\SafeOne_Q(X)$ is the set of configurations for which the state is in $Q$ if the configuration is terminal, and
-every next configuration is in $X$:
+\begin{definition} We characterise partial/total correctness (with
+respect to postcondition $Q$) as the greatest/least fixed point of the
+function $\SafeOne_Q$, defined as follows:
 \[
@@
 \]
+\end{definition}
+%
+We now establish a few properties of $\SafeOne_Q$. The first enables
+the use of the greatest \emph{post}-fixed point and the least
+\emph{pre}-fixed point of $\SafeOne_Q$ as proxies, respectively, for
+its greatest and least fixed points. 
+%
+\begin{lemma}[Monotonicity]
+\label{lem:mono}
+%
+$\SafeOne_Q$ is monotone; i.e., $X \subseteq X'$ implies
+$\SafeOne_Q(X)\subseteq\SafeOne_Q(X')$.
+\end{lemma}
+%
+The next lemmas allow $\SafeOne_Q$'s fixed points to be constructed
+via series of approximants.
+%
+\begin{lemma}[GLB-preservation]
+\label{lem:glb-pres}
+%
+$\SafeOne_Q$ preserves greatest lower bounds. That is, for any
+ascending chain $x_0\subseteq x_1 \subseteq \dots$, we have
+$\SafeOne_Q(\bigcap_k x_k) = \bigcap_k \SafeOne_Q(x_k)$.
+%
+\end{lemma}
+%
+\begin{lemma}[LUB-preservation]
+\label{lem:lub-pres}
+%
+If our transition system has finite branching, then $\SafeOne_Q$
+preserves least upper bounds. That is, if $\finite(\Next(C))$ holds
+for all $C$, then for any ascending chain
+$x_0\subseteq x_1 \subseteq \dots$, we have
+$\SafeOne_Q(\bigcap_k x_k) = \bigcap_k \SafeOne_Q(x_k)$.
+%
+\end{lemma}
+%
+The following lemma states that in the absence of infinite branching,
+every always-terminating command has a longest trace.
+%
+\begin{lemma}[Longest trace]
+\label{lem:longest-trace}
+%
+If $\Next(C)$ is finite for all $C$, and $C_0 ∈ \terminates$, then
+there exists an upper bound $M$ for which
+$∀π ∈ \tracesfrom(C_0)．∃j≤M．\length(π) = j$.
+%
+\end{lemma}
+%
+\begin{proof} The lemma holds because $\tracesfrom(C_0)$ is a finite
+set; this observation in turn rests on K\H{o}nig's infinity lemma.
+\end{proof}
+
+\begin{theorem}
+\label{thm:main}
+%
+Partial and total correctness can be characterised as greatest/least
+fixed points. Note that the third implication below relies
+on our transition system having the property of finite branching.
+%
+\begin{eqnarray}
+\label{thm1_pc} 
+\seqspec{P}c{Q} & = & (\{c\}×P) ⊆ νX\ldotp\SafeOne_Q(X) \\
+\label{thm1_tc1} 
+\totalspec{P}c{Q} & ⇒ & (\{c\}×P) ⊆ μX\ldotp\SafeOne_Q(X) \\
+\label{thm1_tc2}
+\totalspec{P}c{Q} & ~⇐~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X)\quad
+\text{if $\finite(\Next(C))$ for all $C$.}
+\end{eqnarray}
+%
 \end{theorem}
@@
 \begin{proof}
-We begin with the fixed point characterisation of partial correctness. To prove \eqref{thm1_pc}, it suffices to prove
+We begin with the fixed point characterisation of partial
+correctness. To prove \eqref{thm1_pc}, it suffices to prove that
+$\Safe_Q$ coincides with $\SafeOne_Q$'s greatest post-fixed point.
+\begin{eqnarray}
+\label{eq:lem_pc}
+\Safe_Q &=& νX\ldotp \SafeOne_Q(X).
+\end{eqnarray}
+%
+For the $(⊆)$ direction of~\eqref{eq:lem_pc}, it is straightforward to
+show, by expanding definitions and invoking standard lemmas about
+reflexive transitive closures, that $\Safe_Q$ is a post-fixed point
+(that is, $\Safe_Q ⊆ \SafeOne_Q (\Safe_Q)$), and hence that it is
+below the \emph{greatest} post-fixed point. To show the $(⊇)$
+direction, we first observe that we can invoke Lemma~\ref{lem:glb-pres}
+to construct $\SafeOne_Q$'s greatest post-fixed point as the
+intersection of a sequence of approximants:
+%
 \begin{eqnarray*}
-\Safe_Q &=& νX\ldotp \SafeOne_Q(X).
+νX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcap_k\left(\SafeOne_Q^k(\Config)\right).
 \end{eqnarray*}
-For the $(⊆)$ direction, we show that $\Safe_Q$ is a post-fixed point (that is, $\Safe_Q ⊆ \SafeOne_Q (\Safe_Q)$), and immediately deduce that it is below the \emph{greatest} post-fixed point.
+(The intuition is that the $k$th approximant only contains
+configurations for which every trace that terminates in fewer than $k$
+steps satisfies the postcondition when it does so.) It then suffices to
+prove that for all $k$ we have
+\[
+∀C ∈ \SafeOne_Q^k (\Config). ∀π ∈ \tracesfrom(C)．∀j<k．\length(π) = j+1 ⇒ \snd (π_j) ∈ Q,
+\] 
+a fact that can be dispatched via mathematical induction on $k$.
 
-In showing the $(⊇)$ direction, we observe that since $\SafeOne_Q$ preserves greatest lower bounds, the greatest fixed point can be constructed as the intersection of a series of approximants:
+We now turn to the fixed point characterisation of total correctness,
+which is proved using roughly symmetric reasoning. To
+prove~\eqref{thm1_tc1}, it suffices to show
+that
 \begin{eqnarray*}
-νX\ldotp \SafeOne_Q(X) &=& \bigcap_{n∈\nat}\left(\SafeOne_Q^n(\Config)\right).
+\Safe_Q ∩ \terminates &⊇& μX\ldotp \SafeOne_Q(X),
 \end{eqnarray*}
-
-The intuition is that the $k$th approximant only contains configurations for which every trace that terminates in fewer than $k$ steps satisfies the postcondition when it does so.
-
-We now turn to the fixed point characterisation of total correctness. To prove \eqref{thm1_tc}, it suffices to prove
+%
+which follows from $\Safe_Q ∩ \terminates$ being a pre-fixed point of
+$\SafeOne_Q$ and hence above its \emph{least} pre-fixed point. To
+show~\eqref{thm1_tc2}, we first use Lemma~\ref{lem:lub-pres} to equate
+the least pre-fixed point to the union of a sequence of approximants
+as follows:
+%
+\begin{eqnarray*}
+μX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcup_n\left(\SafeOne_Q^n(∅)\right).
+\end{eqnarray*}
+%
+(The intuition is that the $k$th approximant contains all
+configurations for which every trace terminates in fewer than $k$
+steps and satisfies the postcondition when it does so.) We then show,
+by mathematical induction on $k$ as before, that
+%
 \[
-∀C ∈ \SafeOne_Q^k (\Config). ∀π ∈ \tracesfrom(C)．∀j<k．\length(π) = j+1 ⇒ \snd (π_j) ∈ Q
+∀C．(∀π ∈ \tracesfrom(C)． ∃j < k．\length(π) = j+1 ∧ \snd (π_j) ∈ Q) ⇒ C ∈ \SafeOne_Q^{k+1}(\emptyset),
 \]
-which can be dispatched via mathematical induction on $k$.
-
-It suffices to show that
-\begin{eqnarray*}
-\Safe_Q ∩ \terminates &=& μX\ldotp \SafeOne_Q(X).
-\end{eqnarray*}
-For the $(⊇)$ direction, we show that $\Safe_Q$ is a pre-fixed point (that is, $\SafeOne_Q (\Safe_Q) ⊆ \Safe_Q$), and immediately deduce that it is above the \emph{least} pre-fixed point.
-
-In showing the $(⊆)$ direction, we observe that since $\SafeOne_Q$ preserves least upper bounds (this relies on the finitary nature of $\Next(C)$), the least fixed point can be constructed as the union of a series of approximants:
-\begin{eqnarray*}
-μX\ldotp \SafeOne_Q(X) &=& \bigcup_{n∈\nat}\left(\SafeOne_Q^n(∅)\right).
-\end{eqnarray*}
-The intuition is that the $k$th approximant contains all configurations for which every trace terminates in fewer than $k$ steps and satisfies the postcondition when it does so.
-
-It remains then to show
-\[
-∀C．(∀π ∈ \tracesfrom(C)． ∃j < k．\length(π) = j+1 ∧ \snd (π_j) ∈ Q) ⇒ C ∈ \SafeOne_Q^{k+1}(\emptyset)
-\]
-which can be dispatched via mathematical induction on $k$.
+which completes the proof. \qed
 \end{proof}
@@
 \section{Proof mechanisation}
+\label{sec:isabelle}
 
-The proof presented in the previous section has been formalised in the Isabelle theorem prover, with the exception of the following lemma, which has currently only been proved `by hand'.
-It states that in the absence of infinite branching, every command that always terminates has a longest trace.
-\begin{lemma}
-If $\Next(C)$ is finite for all $C$, and $C_0 ∈ \terminates$, then there exists an upper bound $M$ for which $∀π ∈ \tracesfrom(C_0)．∃j≤M．\length(π) = j$.
-\end{lemma}
-The lemma holds because $\tracesfrom(C_0)$ is a finite set; this observation in turn rests on K\H{o}nig's infinity lemma.
+The proof presented in the previous section has been formalised in the
+Isabelle theorem prover, with the exception of the following lemma,
+which has currently only been proved `by hand'.
 
 \section{On infinite branching}
+\label{sec:infinite_branching}
 
-If the $\trans$ relation allows infinite branching -- that is, if a configuration can have infinitely many next configurations to choose from -- then the least fixed point calculation does not coincide with total correctness.
-The technical reason for the failure is that $\SafeOne_Q$ no longer preserves least upper bounds.
-Intuitively, the failure can be explained by the following counterexample.
+If the $\trans$ relation allows infinite branching -- that is, if a
+configuration can have infinitely many next configurations to choose
+from -- then the least fixed point calculation does not coincide with
+total correctness. The technical reason for the failure is that
+$\SafeOne_Q$ no longer preserves least upper bounds. Intuitively, the
+failure can be explained by the following counterexample.
 \[
@@
 }] (c43) -- (c44);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] [decorate,decoration={brace,amplitude=10pt}]
+\draw[decorate,decoration={brace,amplitude=10pt}]
 ([xshift=5mm]c44.north) -- ([xshift=5mm]c42.south) node [black,midway,xshift=6mm] {$k$};
@@
 \]
-The $k$th approximant contains configurations whose traces all terminate in fewer than $k$ steps.
-No approximant can contain configuration $C_0$, since there is no bound $k$ on the length of its traces.
-Yet this configuration will be admitted by a total correctness specification whose postcondition is $\mathit{true}$, since it is the case that each of its traces terminates.
+The $k$th approximant contains configurations whose traces all
+terminate in fewer than $k$ steps. No approximant can contain
+configuration $C_0$, since there is no bound $k$ on the length of its
+traces. Yet this configuration will be admitted by a total correctness
+specification whose postcondition is $\mathit{true}$, since it is the
+case that each of its traces terminates.
 
-Therefore, in the presence of infinite branching, the fixed point characterisations of total correctness must be weakened to an implication.
-Preservation of greatest lower bounds is unaffected by infinite branching, so the characterisation of partial correctness remains intact.
+Therefore, in the presence of infinite branching, the fixed point
+characterisations of total correctness must be weakened to an
+implication. Preservation of greatest lower bounds is unaffected by
+infinite branching, so the characterisation of partial correctness
+remains intact.
 
@@
 
-\section{Discussion}
+\section{Related work}
+\label{sec:related}
 
-Recent work on various Hoare logics for concurrency have used (mildly disguised) greatest fixed point calculations to obtain a partial correctness semantics.
-See, for example, \cite[Definition~3.2]{vafeiadis11} and \cite[Definition~25]{dinsdale-young+13}.
-The result in this paper could prove helpful when those program logics are extended to total correctness.
+Clarke~\cite{clarke79} was the first to study the correspondence
+between greatest/least fixed points and partial/total correctness. He
+remarks that
+\begin{quote}the fixedpoints of $Γ$ form a complete lattice under the
+natural partial ordering on $\pow(\State)$. The top element of this
+lattice is the weakest precondition for partial correctness and the
+bottom element is the weakest precondition for total
+correctness.~\cite[p.~279]{clarke79}
+\end{quote} 
+%
+We note three ways in which the current work extends on Clarke's
+original observation. First, Clarke's work is in the setting of a
+particular programming language that provides syntax for sequencing,
+conditionals, assignment and recursive calls to parameterless
+procedures. In this paper, we need not tackle issues of programming
+language syntax, since we work at the level of arbitrary transition
+systems. Second, Clarke's work is in the setting of a big-step
+semantics; that is, the meaning of each programming construct maps an
+initial state directly to a final state. We work with a small-step
+semantics (as described by a transition system), which means that our
+approach handles parallel programs without any further work. Big-step
+semantics is well-known to be unable to handle parallelism, as
+acknowledged by Clarke, who remarks that he `[is] currently attempting
+to extend this fixedpoint theory to additional programming features
+such as parallelism'~\cite[p.~292]{clarke79}. The definition of total
+correctness in a small-step setting requires more nuance than in a
+big-step setting, where it is trivial. Third, Clarke's work applies
+only in a deterministic setting: a program has either a single final
+state, or none at all. We allow for highly non-deterministic programs,
+and only restrict the amount of non-determinism in the case of total
+correctness.
 
-Jacobs and Gries have proposed \emph{general correctness} as a way to unify partial and total correctness~\cite{jacobs+85}.
-It would be interesting to investigate whether general correctness can be also characterised as a fixed point calculation.
+Recent work on various Hoare logics for concurrency have used (mildly
+disguised) greatest fixed point calculations to obtain a partial
+correctness semantics. See, for example,
+\cite[Definition~3.2]{vafeiadis11} and
+\cite[Definition~25]{dinsdale-young+13}. The result in this paper
+could prove helpful when those program logics are extended to total
+correctness.
 
-The idea of using the least and greatest fixed point of the same function has been previously exploited by Paulson~\cite[\S3]{paulson97a}, who obtains the set of finite lists from a function's least fixed point, and the set of possibly-infinite (`lazy') lists from its greatest fixed point.
+Jacobs and Gries have proposed \emph{general correctness} as a way to
+unify partial and total correctness~\cite{jacobs+85}. It would be
+interesting to investigate whether general correctness can be also
+characterised as a fixed point calculation.
 
-\paragraph{Acknowledgements} This work was supported by EPSRC grant EP/K011499/1. I would like to thank Andreas Lochbihler for several helpful discussions.
+The idea of using the least and greatest fixed point of the same
+function has been previously exploited by
+Paulson~\cite[\S3]{paulson97a}, who obtains the set of finite lists
+from a function's least fixed point, and the set of possibly-infinite
+(`lazy') lists from its greatest fixed point.
+
+\paragraph{Acknowledgements} This work was supported by EPSRC grant
+EP/K011499/1. I would like to thank Edmund Clarke, Alastair Donaldson,
+Tony Hoare, Peter Lammich, and Andreas Lochbihler for helpful feedback
+and discussions.
 
 \bibliographystyle{abbrv}
-\bibliography{../../John}
+\bibliography{/Users/jpw48/Dropbox/John}
 

### 2015-01-21 paper.tex
+

@@
 \usepackage{bm}
+\usepackage{framed}
 
@@
 \usepackage[utf8x]{inputenc}
+\usepackage{url}
 
@@
 total correctness, that differ only in that the former takes a
-function's greatest fixed point where the latter takes its least, and
-Section~\ref{sec:isabelle} describes how this characterisation is
-proved correct using the Isabelle theorem
-prover. Section~\ref{sec:infinite_branching} describes a condition
-that is necessary for the fixed point characterisation of \emph{total}
+function's greatest fixed point where the latter takes its least.
+Section~\ref{sec:infinite_branching} describes a condition that is
+necessary for the fixed point characterisation of \emph{total}
 correctness to be accurate; namely, that there is no infinite
-branching. We begin, in Section~\ref{sec:preliminaries}, by
-establishing some preliminary definitions.
+branching. Section~\ref{sec:related} discusses related work, mainly
+focussing on how our result extends a similar observation made by
+Clarke in 1979~\cite{clarke79}. We begin, in
+Section~\ref{sec:preliminaries}, by establishing some preliminary
+definitions.
 
@@
 \[\begin{array}{r@{~~}c@{~~}l}
-\BOX p & = & \{C \mid ∀C'∈\Next(C)\ldotp C' ∈ p\} \\
-\DIAMOND p & = & \{C \mid ∃C'∈\Next(C)\ldotp C' ∈ p\} \\
-μX\ldotp \varphi(X) & = & \bigcap\{S\mid \varphi(S) ⊆ S\} \\
-νX\ldotp \varphi(X) & = & \bigcup\{S\mid S ⊆ \varphi(S)\}
+\BOX p & = & \{C．∀C'∈\Next(C)\ldotp C' ∈ p\} \\
+\DIAMOND p & = & \{C． ∃C'∈\Next(C)\ldotp C' ∈ p\} \\
+μX\ldotp \varphi(X) & = & \bigcap\{S．\varphi(S) ⊆ S\} \\
+νX\ldotp \varphi(X) & = & \bigcup\{S．S ⊆ \varphi(S)\}
 \end{array}\]
@@
 
-\begin{definition} We characterise partial/total correctness (with
+\begin{definition}
+\label{defn:safeone}
+We characterise partial/total correctness (with
 respect to postcondition $Q$) as the greatest/least fixed point of the
@@
 \begin{array}{r@{~~}c@{~~}l}
-\SafeOne_Q(X) &\eqdef& \{\config{c}{σ}\mid  \stuck{\config{c}{σ}} ⇒ σ ∈ Q\} ∩ \Box X.
+\SafeOne_Q(X) &\eqdef& \{\config{c}{σ}．\stuck{\config{c}{σ}} ⇒ σ ∈ Q\} ∩ \Box X.
 \end{array}
@@
 %
-\begin{proof} The lemma holds because $\tracesfrom(C_0)$ is a finite
-set; this observation in turn rests on K\H{o}nig's infinity lemma.
+\begin{proof} We first recall K\H{o}nig's infinity lemma as it applies
+to trees -- that every tree with infinitely-many vertices, each having
+finitely-many successor vertices, has at least one infinite
+trace~\cite{franchella97}:
+\begin{equation}
+\label{eq:konig}
+¬\finite\{C'．C_0 \trans^* C'\} ⇒ (∀C．\finite(\Next(C))) ⇒
+∃π∈\tracesfrom(C_0)．\dom(π) = \nat.
+\end{equation}
+Now, if $C_0∈\terminates$, then it has no infinite traces, and
+hence, by the contrapositive of~\eqref{eq:konig}, the number
+of configurations reachable from $C_0$ is finite. This number provides
+a suitable upper bound $M$ on the length of traces from $C_0$. 
 \end{proof}
-
+%
+\begin{framed}
 \begin{theorem}
@@
 \label{thm1_tc1} 
-\totalspec{P}c{Q} & ⇒ & (\{c\}×P) ⊆ μX\ldotp\SafeOne_Q(X) \\
+\totalspec{P}c{Q} & ⇐ & (\{c\}×P) ⊆ μX\ldotp\SafeOne_Q(X) \\
 \label{thm1_tc2}
-\totalspec{P}c{Q} & ~⇐~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X)\quad
+\totalspec{P}c{Q} & ~⇒~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X)\quad
 \text{if $\finite(\Next(C))$ for all $C$.}
@@
 \end{theorem}
-
+\end{framed}
+%
 \begin{proof}
@@
 below the \emph{greatest} post-fixed point. To show the $(⊇)$
-direction, we first observe that we can invoke Lemma~\ref{lem:glb-pres}
-to construct $\SafeOne_Q$'s greatest post-fixed point as the
-intersection of a sequence of approximants:
+direction, we first observe that we can invoke
+Lemma~\ref{lem:glb-pres} to construct $\SafeOne_Q$'s greatest
+post-fixed point as the intersection of a sequence of approximants:
 %
@@
 \end{eqnarray*}
-(The intuition is that the $k$th approximant only contains
-configurations for which every trace that terminates in fewer than $k$
-steps satisfies the postcondition when it does so.) It then suffices to
-prove that for all $k$ we have
+(The intuition is that approximant $k+1$ contains configurations for
+which every stuck configuration that is reachable in $k$ steps
+satisfies the postcondition.) After expanding the definition of
+$\Safe$, it remains to show:
+%
+\begin{equation}
+\label{eq:pc_mid}
+(∀k．C ∈ \SafeOne_Q^k(\Config)) ⇒ ∀C'．((C \trans^* C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q.
+\end{equation}
+%
+After rewriting $C \trans^* C'$ as $∃n．C \trans^n C'$ and then
+instantiating the $k$ in~\eqref{eq:pc_mid} to $n+1$, it suffices to show, for all $n$:
 \[
-∀C ∈ \SafeOne_Q^k (\Config). ∀π ∈ \tracesfrom(C)．∀j<k．\length(π) = j+1 ⇒ \snd (π_j) ∈ Q,
+C ∈ \SafeOne_Q^{n+1}(\Config) ⇒ ∀C'．((C \trans^n C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q,
 \]
-a fact that can be dispatched via mathematical induction on $k$.
+which can be dispatched via mathematical induction on $n$ (with
+$C$ universally quantified in the induction hypothesis).
 
-We now turn to the fixed point characterisation of total correctness,
-which is proved using roughly symmetric reasoning. To
+We now turn to the fixed point characterisation of total correctness. To
 prove~\eqref{thm1_tc1}, it suffices to show
@@
 \begin{eqnarray*}
-μX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcup_n\left(\SafeOne_Q^n(∅)\right).
+μX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcup_k\left(\SafeOne_Q^k(∅)\right).
 \end{eqnarray*}
 %
-(The intuition is that the $k$th approximant contains all
-configurations for which every trace terminates in fewer than $k$
-steps and satisfies the postcondition when it does so.) We then show,
-by mathematical induction on $k$ as before, that
+(The intuition is that approximant $k+1$ contains all configurations
+for which every trace terminates in no more than $k$ steps and satisfies
+the postcondition when it does so.) After expanding the definition of
+$\Safe$, and invoking Lemma~\ref{lem:longest-trace} to obtain the
+length $M$ of $C$'s longest trace, it remains to show:
+%
+\begin{equation}
+\label{eq:tc_mid}
+\begin{array}{l}
+(∃M．∀π ∈ \tracesfrom(C)．∃j≤M．\length(π) = j) ⇒ {}\\
+(∀C'．((C \trans^* C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q) ⇒ (∃k．C ∈ \SafeOne_Q^k(\emptyset)).
+\end{array}
+\end{equation}
+%
+We choose $M$ as a witness for $k$ in~\eqref{eq:tc_mid}. It then
+suffices to show, for all $M$:
 %
 \[
-∀C．(∀π ∈ \tracesfrom(C)． ∃j < k．\length(π) = j+1 ∧ \snd (π_j) ∈ Q) ⇒ C ∈ \SafeOne_Q^{k+1}(\emptyset),
+\begin{array}{l}
+(∀π ∈ \tracesfrom(C)．∃j≤M．\length(π) = j) ⇒ {}\\
+(∀C'．((C \trans^* C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q) ⇒ C ∈ \SafeOne_Q^M(\emptyset).
+\end{array}
 \]
-which completes the proof. \qed
+%
+With a little further algebraic manipulation we arrive at:
+\[
+(∀π ∈ \tracesfrom(C)．∃j≤M．\length(π) = j ∧ \snd(\pi_{j-1}) ∈ Q) ⇒ C ∈ \SafeOne_Q^M(\emptyset)
+\]
+which can be dispatched by mathematical induction on $M$, generalising
+$C$ as before, thus completing the proof.\qed
 \end{proof}
 
-\section{Proof mechanisation}
-\label{sec:isabelle}
-
-The proof presented in the previous section has been formalised in the
-Isabelle theorem prover, with the exception of the following lemma,
-which has currently only been proved `by hand'.
+\paragraph{Proof mechanisation} Our main theorem and its accompanying
+lemmas have been formalised and proved in the Isabelle theorem prover,
+with the exception of Lemma~\ref{lem:longest-trace}, of which we
+currently only have a hand proof. The proof relies on K\H{o}nig's
+infinity lemma, which we found to be difficult to formalise. The
+proof is available online.\footnote{\url{http://www.doc.ic.ac.uk/~jpw48/Partial_Total_Correctness_As_Fixed_Points.thy}}
 
@@
 total correctness. The technical reason for the failure is that
-$\SafeOne_Q$ no longer preserves least upper bounds. Intuitively, the
+$\SafeOne_Q$ no longer preserves least upper bounds
+(cf.~Lemma~\ref{lem:lub-pres}). Intuitively, the
 failure can be explained by the following counterexample.
+%
 \[
-\begin{tikzpicture}
+\begin{tikzpicture}[zaggy/.style={line join=round,
+decorate, ->,decoration={zigzag, segment length=4,
+amplitude=.9,post=lineto, post length=2pt}}]
 \node[draw] (c0) at (3,0) {$C_0$};
@@
 \node             at (6,1) {$\dots$};
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c0) -- (c21);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c21) -- (c22);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c0) -- (c31);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c31) -- (c32);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c32) -- (c33);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c0) -- (c11);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c11) -- (c12);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c12) -- (c13);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c13) -- (c14);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c0) -- (c41);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c41) -- (c42);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c42) -- (c43);
-\draw[line join=round,
-decorate, ->,decoration={
-    zigzag,
-    segment length=4,
-    amplitude=.9,post=lineto,
-    post length=2pt
-}] (c43) -- (c44);
+\draw[zaggy] (c0) to [in=270,out=180] (c21);
+\draw[zaggy] (c21) -- (c22);
+\draw[zaggy] (c0) to [in=270,out=180] (c31);
+\draw[zaggy] (c31) -- (c32);
+\draw[zaggy] (c32) -- (c33);
+\draw[zaggy] (c0) -- (c11);
+\draw[zaggy] (c11) -- (c12);
+\draw[zaggy] (c12) -- (c13);
+\draw[zaggy] (c13) -- (c14);
+\draw[zaggy] (c0) to[in=270,out=0] (c41);
+\draw[zaggy] (c41) -- (c42);
+\draw[zaggy] (c42) -- (c43);
+\draw[zaggy] (c43) -- (c44);
 \draw[decorate,decoration={brace,amplitude=10pt}]
-([xshift=5mm]c44.north) -- ([xshift=5mm]c42.south) node [black,midway,xshift=6mm] {$k$};
+([xshift=5mm]c44.north) -- ([xshift=5mm]c42.south) node
+[black,midway,xshift=6mm] {$k$};
 \end{tikzpicture}
 \]
-The $k$th approximant contains configurations whose traces all
+Recall that approximant $k$ contains configurations whose traces all
 terminate in fewer than $k$ steps. No approximant can contain
-configuration $C_0$, since there is no bound $k$ on the length of its
+configuration $C_0$, since there is no bound on the length of its
 traces. Yet this configuration will be admitted by a total correctness
@@
 Therefore, in the presence of infinite branching, the fixed point
-characterisations of total correctness must be weakened to an
+characterisation of total correctness must be weakened to an
 implication. Preservation of greatest lower bounds is unaffected by
@@
 remains intact.
-
-\begin{eqnarray*}
-\seqspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ νX\ldotp \SafeOne_Q(X) \\
-\totalspec{P}c{Q} & ~⇐~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X) 
-\end{eqnarray*}
 
@@
 between greatest/least fixed points and partial/total correctness. He
-remarks that
-\begin{quote}the fixedpoints of $Γ$ form a complete lattice under the
-natural partial ordering on $\pow(\State)$. The top element of this
-lattice is the weakest precondition for partial correctness and the
-bottom element is the weakest precondition for total
-correctness.~\cite[p.~279]{clarke79}
-\end{quote} 
+remarks, in a 1979 article, that 
+%
+\begin{quote}the fixedpoints of $\Gamma$ form a complete
+lattice under the natural partial ordering on $\pow(\State)$. The top
+element of this lattice is the weakest precondition for partial
+correctness and the bottom element is the weakest precondition for
+total correctness.~\cite[p.~279, footnote]{clarke79} \end{quote}
 %
 We note three ways in which the current work extends on Clarke's
-original observation. First, Clarke's work is in the setting of a
+original observation. First, Clarke works in the setting of a
 particular programming language that provides syntax for sequencing,
-conditionals, assignment and recursive calls to parameterless
+conditionals, assignment, and recursive calls to parameterless
 procedures. In this paper, we need not tackle issues of programming
 language syntax, since we work at the level of arbitrary transition
-systems. Second, Clarke's work is in the setting of a big-step
-semantics; that is, the meaning of each programming construct maps an
-initial state directly to a final state. We work with a small-step
-semantics (as described by a transition system), which means that our
-approach handles parallel programs without any further work. Big-step
-semantics is well-known to be unable to handle parallelism, as
-acknowledged by Clarke, who remarks that he `[is] currently attempting
-to extend this fixedpoint theory to additional programming features
-such as parallelism'~\cite[p.~292]{clarke79}. The definition of total
-correctness in a small-step setting requires more nuance than in a
-big-step setting, where it is trivial. Third, Clarke's work applies
-only in a deterministic setting: a program has either a single final
-state, or none at all. We allow for highly non-deterministic programs,
-and only restrict the amount of non-determinism in the case of total
-correctness.
+systems. Second, Clarke's work applies only in a deterministic
+setting: a program has either a single final state, or none at all. We
+allow for highly non-deterministic programs, and only restrict to
+finite non-determinism in the case of total correctness. Third,
+Clarke's work is in the setting of a big-step semantics; that is, the
+meaning of each programming construct maps an initial state directly
+to a final state. We work with a small-step semantics (as described by
+a transition system), which means that our approach handles parallel
+programs without further adaptation. Big-step semantics is well-known
+to be unable to handle parallelism, as acknowledged by Clarke, who
+remarks that he `[is] currently attempting to extend this fixedpoint
+theory to additional programming features such as
+parallelism'~\cite[p.~292]{clarke79}. The definition of total
+correctness in a big-step and deterministic setting is trivial, and
+hence Clarke's $\Gamma$ functional is straightforward -- this perhaps
+explains why Clarke consigned the observation we quote above to a mere
+footnote. Our functional, on the other hand, which we call $\SafeOne$,
+is fairly subtle: experience shows that even minor modifications of
+Definition~\ref{defn:safeone} quickly lead to either the least or the
+greatest fixed point becoming degenerate.
 
-Recent work on various Hoare logics for concurrency have used (mildly
-disguised) greatest fixed point calculations to obtain a partial
-correctness semantics. See, for example,
+Regarding other influences on this work: the idea of using the least
+and greatest fixed point of the same function has been previously
+exploited by Paulson~\cite[\S3]{paulson97a}, who obtains the set of
+finite lists from a function's least fixed point, and the set of
+possibly-infinite (`lazy') lists from its greatest fixed point.
+
+Recent work on various Hoare logics for concurrency provide avenues
+for further development of the current work. These logics typically
+use (mildly disguised) greatest fixed point calculations to obtain a
+partial correctness semantics; see, for example,
 \cite[Definition~3.2]{vafeiadis11} and
-\cite[Definition~25]{dinsdale-young+13}. The result in this paper
-could prove helpful when those program logics are extended to total
-correctness.
+\cite[Definition~25]{dinsdale-young+13}. If these logics were extended
+to handle total correctness, the result presented in this paper could
+ease the transition.
 
-Jacobs and Gries have proposed \emph{general correctness} as a way to
-unify partial and total correctness~\cite{jacobs+85}. It would be
-interesting to investigate whether general correctness can be also
-characterised as a fixed point calculation.
-
-The idea of using the least and greatest fixed point of the same
-function has been previously exploited by
-Paulson~\cite[\S3]{paulson97a}, who obtains the set of finite lists
-from a function's least fixed point, and the set of possibly-infinite
-(`lazy') lists from its greatest fixed point.
-
+Another direction for future work is provided by Jacobs and Gries, who
+have proposed \emph{general correctness} as a way to unify partial and
+total correctness~\cite{jacobs+85}. It would be interesting to
+investigate whether general correctness can be also characterised as a
+fixed point calculation.
+\[
+\text{---}
+\]
 \paragraph{Acknowledgements} This work was supported by EPSRC grant
@@
 
+\newpage
 \bibliographystyle{abbrv}

### 2015-01-21 paper.tex
+

@@
 In the context of Hoare Logic~\cite{hoare69}, a (possibly
-non-deterministic, possibly non-terminating) program satisfies a
+non-deterministic, and possibly non-terminating) program satisfies a
 \emph{total correctness} specification when each of its traces from a
@@
 connection between total/partial correctness and least/greatest fixed
-points. This paper seeks to state this connection precisely, and
+points. This paper seeks to state that connection precisely, and
 investigate conditions under which it holds.
@@
 focussing on how our result extends a similar observation made by
-Clarke in 1979~\cite{clarke79}. We begin, in
-Section~\ref{sec:preliminaries}, by establishing some preliminary
-definitions.
+Edmund Clarke in 1979. We begin, in Section~\ref{sec:preliminaries},
+by establishing some preliminary definitions.
 
@@
 \paragraph{Proof mechanisation} Our main theorem and its accompanying
-lemmas have been formalised and proved in the Isabelle theorem prover,
-with the exception of Lemma~\ref{lem:longest-trace}, of which we
-currently only have a hand proof. The proof relies on K\H{o}nig's
-infinity lemma, which we found to be difficult to formalise. The
-proof is available online.\footnote{\url{http://www.doc.ic.ac.uk/~jpw48/Partial_Total_Correctness_As_Fixed_Points.thy}}
+lemmas have been formalised and proved in the Isabelle theorem
+prover~\cite{nipkow+02}, with the exception of
+Lemma~\ref{lem:longest-trace}, of which we currently only have a hand
+proof. The proof relies on K\H{o}nig's infinity lemma, which we found
+to be difficult to formalise. The proof is available
+online.\footnote{\url{http://www.doc.ic.ac.uk/~jpw48/Partial_Total_Correctness_As_Fixed_Points.thy}}
 
@@
 
-Clarke~\cite{clarke79} was the first to study the correspondence
-between greatest/least fixed points and partial/total correctness. He
-remarks, in a 1979 article, that 
+Edmund Clarke~\cite{clarke79} was probably the first to study the
+correspondence between greatest/least fixed points and partial/total
+correctness. He remarks, in a 1979 article, that
 %
